Deep mutational scanning of hemagglutinin helps distinguish the evolutionary fates of human H3N2 influenza virus lineages

### 2018-01-23 paper/manuscript.tex
Change color of sialic acid receptor in mut_tolerance figure
Small tweak to dms_overview fig, update phydms table

\subsection*{Our measurements are consistent with existing knowledge about HA's evolution and function}
How do the HA amino-acid preferences measured in our experiments relate to the evolution of H3N2 influenza virus in nature?
This question can be addressed by evaluating how well an experimentally informed codon substitution model (ExpCM) using our measurements describe H3N2 evolution compared to standard phylogenetic substitution models~\cite{bloom2017identification,hilton2017phydms}.
Table~\ref{tab:phydms} shows that the ExpCM using the across-replicate average of our measurements greatly outperforms conventional substitution models.
This result indicates that our experiments authentically capture some of the constraints on HA evolution. 
The relative rate of nonsynonymous to synonymous substitutions (dN/dS or $\omega$) is $\ll$1 for conventional substitution models (Table~\ref{tab:phydms}).
However, the relative rate of nonsynonymous to synonymous substitutions after accounting for the amino-acid constraints measured in our experiments (the $\omega$ for the ExpCM) is close to one (Table~\ref{tab:phydms}), indicating that the deep mutational scanning captures much of the purifying selection on HA.
ExpCM fit a stringency parameter that relates the selection in the experiments to that in nature~\cite{bloom2017identification,hilton2017phydms}.
The stringency parameter for our HA ExpCM is 2.44 (Table~\ref{tab:phydms}), indicating that natural selection favors the same amino acids as the experiments, but with greater stringency.
Throughout the rest of this paper, we use experimental measurements re-scaled~\cite{bloom2017identification,hilton2017phydms} by this stringency parameter.
These re-scaled preferences are shown in Figure~\ref{fig:logoplot}.

\begin{table}
\caption{\label{tab:phydms}
{\bf Substitution models informed by the experiments describe H3N2's natural evolution better than traditional substitution models.}}
\begin{center} 
\begin{tabular}{cccccccc}
\hline
\bf{Model} & \bf{$\Delta$AIC} & \bf{LnL} & \bf{Stringency} & \bf{$\omega$}  \\ \hline
ExpCM & 0.0 & -8439 & 2.44 & 0.91 \\
GY94 M5 & 2166 & -9516 & -- & 0.36 (0.30, 0.84) \\
ExpCM, site avg. & 2504 & -9691 & 0.68 & 0.32 \\
GY94 M0 & 2608 & -9738 & -- & 0.31 \\
\hline
\end{tabular}
 \end{center}
\addtabletext{Maximum likelihood phylogenetic fit to an alignment of human H3N2 influenza HAs using ExpCM~\cite{hilton2017phydms}, ExpCM in which the experimental measurements are averaged across sites (site avg.), and the M0 and M5 versions of the Goldman-Yang (GY94) model~\cite{yang2000codon}.
Models are compared by AIC~\cite{posada2004model} computed from the log likelihood (LnL) and number of model parameters.
The $\omega$ parameter is dN/dS for the Goldman-Yang models, and the relative dN/dS after accounting for the measurements for the ExpCM.
For the M5 model, we give the mean followed by the shape and rate parameters of the gamma distribution over $\omega$.
}
\end{table}

@@

%\begin{SCfigure*}[\sidecaptionrelwidth][t]
\begin{figure}
\centering
%\includegraphics[width=11.4cm]{figs/mut_tolerance/entropy_heatmap.pdf}
\includegraphics[width=\columnwidth]{figs/mut_tolerance/entropy_heatmap.pdf}
\caption{\label{fig:mut_tolerance}
{\bf Mutational tolerance of each site in H3 and H1 HAs.}
Mutational tolerance as measured in the current study is mapped onto the structure of the H3 trimer (PDB 4O5N;~\cite{lee2014receptor}).
Mutational tolerance of the WSN/1933 H1 HA as measured in \cite{doud2016accurate} is mapped onto the structure of the H1 trimer (PDB 1RVX;~\cite{gamblin2004structure}).
Different color scales are used because measurements are comparable among sites within the same HA, but not necessarily across HAs.
Both trimers are shown in approximately the same orientation. 
For each HA, the structure at left shows a surface representation of the full trimer, while the structure at right side shows a ribbon representation of just one monomer.
The sialic acid receptor is shown in black sticks.
}
%\end{SCfigure*}
\end{figure}

>>>>

\subsection*{Our measurements are consistent with existing knowledge about HA's evolution and function}
How do the HA amino-acid preferences measured in our experiments relate to the evolution of H3N2 influenza virus in nature?
This question can be addressed by evaluating how well an experimentally informed codon substitution model (ExpCM) using our measurements describe H3N2 evolution compared to standard phylogenetic substitution models~\cite{bloom2017identification,hilton2017phydms}.
Table~\ref{tab:phydms} shows that the ExpCM using the across-replicate average of our measurements greatly outperforms conventional substitution models.
This result indicates that our experiments authentically capture some of the constraints on HA evolution. 
The relative rate of nonsynonymous to synonymous substitutions (dN/dS or $\omega$) is $\ll$1 for conventional substitution models (Table~\ref{tab:phydms}).
However, the relative rate of nonsynonymous to synonymous substitutions after accounting for the amino-acid constraints measured in our experiments (the $\omega$ for the ExpCM) is close to one (Table~\ref{tab:phydms}), indicating that the deep mutational scanning captures much of the purifying selection on HA.
ExpCM fit a stringency parameter that relates the selection in the experiments to that in nature~\cite{bloom2017identification,hilton2017phydms}.
The stringency parameter for our HA ExpCM is 2.47 (Table~\ref{tab:phydms}), indicating that natural selection favors the same amino acids as the experiments, but with greater stringency.
Throughout the rest of this paper, we use experimental measurements re-scaled~\cite{bloom2017identification,hilton2017phydms} by this stringency parameter.
These re-scaled preferences are shown in Figure~\ref{fig:logoplot}.

\begin{table}
\caption{\label{tab:phydms}
{\bf Substitution models informed by the experiments describe H3N2's natural evolution better than traditional substitution models.}}
\begin{center} 
\begin{tabular}{cccccccc}
\hline
\bf{Model} & \bf{$\Delta$AIC} & \bf{LnL} & \bf{Stringency} & \bf{$\omega$}  \\ \hline
ExpCM & 0.0 & -8441 & 2.47 & 0.91 \\
GY94 M5 & 2162 & -9516 & -- & 0.36 (0.30, 0.84) \\
ExpCM, site avg. & 2501 & -9692 & 0.67 & 0.32 \\
GY94 M0 & 2604 & -9738 & -- & 0.31 \\
\hline
\end{tabular}
 \end{center}
\addtabletext{Maximum likelihood phylogenetic fit to an alignment of human H3N2 influenza HAs using ExpCM~\cite{hilton2017phydms}, ExpCM in which the experimental measurements are averaged across sites (site avg.), and the M0 and M5 versions of the Goldman-Yang (GY94) model~\cite{yang2000codon}.
Models are compared by AIC~\cite{posada2004model} computed from the log likelihood (LnL) and number of model parameters.
The $\omega$ parameter is dN/dS for the Goldman-Yang models, and the relative dN/dS after accounting for the measurements for the ExpCM.
For the M5 model, we give the mean followed by the shape and rate parameters of the gamma distribution over $\omega$.
}
\end{table}

@@

%\begin{SCfigure*}[\sidecaptionrelwidth][t]
\begin{figure}
\centering
%\includegraphics[width=11.4cm]{figs/mut_tolerance/entropy_heatmap.pdf}
\includegraphics[width=\columnwidth]{figs/mut_tolerance/entropy_heatmap.pdf}
\caption{\label{fig:mut_tolerance}
{\bf Mutational tolerance of each site in H3 and H1 HAs.}
Mutational tolerance as measured in the current study is mapped onto the structure of the H3 trimer (PDB 4O5N;~\cite{lee2014receptor}).
Mutational tolerance of the WSN/1933 H1 HA as measured in \cite{doud2016accurate} is mapped onto the structure of the H1 trimer (PDB 1RVX;~\cite{gamblin2004structure}).
Different color scales are used because measurements are comparable among sites within the same HA, but not necessarily across HAs.
Both trimers are shown in approximately the same orientation. 
For each HA, the structure at left shows a surface representation of the full trimer, while the structure at right side shows a ribbon representation of just one monomer.
The sialic acid receptor is shown in red sticks.
}
%\end{SCfigure*}
\end{figure}
